An algorithm for finding the global extremum of a partially defined function\thanks{Supported by organization x.}

### 2023-08-27 2023/MMST/samplepaper.tex
fix titles and itemize

\section{Problem Statement}
In the general form, the global optimization problem can be formulated as follows:
\begin{equation}\label{eq1} 
\phi^*=\phi(y^* )=\min_{y \in D} \phi(y), D=\left\{ y \in R^N: a_i \leq y_i \leq b_i, a,b \in R^N,1 \leq i \leq N \right\},
\end{equation}
where $y=(y_1,y_2,...,y_N)$ - is the vector of the varied parameters, $D$ - is an $N$-dimensional hypercube, and $N$ –  is the dimension of the problem being solved.

@@

From our experience in solving the applied optimization problems [...refs??...], we will assume the total volume of the non-computability regions to be a small fraction of the whole volume of the search domain $D$ (about 10-20\%)\footnote{When testing the algorithm, we have considered also more complex problem statements, which the volume of the non-computability region achieved 50-80\% in. The problems of this kind were solved by the algorithm proposed successfully as well but a large number of trials was required to solve the problem because the method attempted to explore the whole non-computability region. In this connection, in the rules of constructing the problems the experiments for the dimensionality $N=5$ were modified so that the non-computability region were located in the corners of the search domain only.}.

@@

\subsubsection{Global search algorithm (GSA).} At every iteration of the global search a \textit{search trial} is executed. We will call a trial a computing the optimized function value  $\phi (x)$ from (\ref{eq2}). The first trial will be performed in the middle internal point $x^1 \in (0,1)$. The choice of the point $x^{k+1}, k \leq 1$ of the next $(k+1)^\text{th}$ trial is performed on the base of the following rules.

\paragraph{Rule 1.} Renumber (by the lower indices) the points $x^i, 1 \leq i \leq k$  of the previous trials in the ascending order by their coordinates i. e.
\begin{equation}\label{eq5} 
0=x_0 < x_1 < ... < x_i < ... < x_k=1
\end{equation}
and juxtapose these ones to the values $z_i=\phi (x_i), 1 \leq i \leq k$ computed in these points and to the indices $v_i=v(x_i)$ defined by the rule
\begin{equation}\label{eq6} 
v_i=v(x_i)=
  \begin{cases}
    -1 & {\quad \text{if } x_i \text{ is a boundary point,}}\\
    1  & {\quad \text{if } x_i \text{ is an internal point,}}
  \end{cases}
\end{equation}
the points $x_0=0$ and $x_{k+1}=1$ are introduced additionally (the values $z_0$ and $z_{k+1}$ are undefined) for convenience of the following notations.

\paragraph{Rule 2.} Compute current lower estimate
\begin{equation}\label{eq7} 
\mu = \max\left\{ \frac{|z_i-z_{i-1}|}{\Delta _i}, 2 \leq i \leq k \right\} , \Delta _i= \| x_i -x_{i-1} \| =(x_i-x_{i-1})^{1/N},
\end{equation}
of the unknown H{\"o}lder constant $K$ from (\ref{eq4}) for the reduced function $\phi(x)$ where $r, r>1$ is a parameter of the algorithm; if $\mu=0$ accept $\mu=1$.
\paragraph{Rule 3.} Determine the best current value of the objective function
\begin{equation}\label{eq8} 
z^*=\min \left\{ \phi (x_i): 1\leq i \leq k \right\}.
\end{equation}

\paragraph{Rule 4.} For each interval $(x_{i-1},x_i),1 \leq i \leq k+1$ compute the value $R(i)$ called a \textit{characteristic} of the interval according to the formula
\begin{equation}\label{eq9} 
R(x_i)=
  \begin{cases}
    \Delta _i+\frac {{(z_i-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-1}-2z^*}{r \mu}, & {\quad  v(x_i)=v(x_{i-1})=1}\\
    2 \Delta _i-4 \frac {(z_i-z^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=1}\\
    2 \Delta _i-4 \frac {(z_{i-1}-z^*)}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=-1}
  \end{cases}
\end{equation}

\paragraph{Rule 5.} Select the interval $t$ with the largest value of characteristic $R(i):R(t)= \max_{1 \leq i \leq k+1} R(i)$.

\paragraph{Rule 6.} Perform the next trial in the middle point of the interval $(x_{t-1},x_t)$ if the indices of its end points are not the same i. e.
\begin{equation}\label{eq10} 
x^{k+1}=\frac {x_t+x_{t-1}}{2}, v(x_{t-1})\neq v(x_t),
\end{equation}
Otherwise, perform the next trial in the point
\begin{equation}\label{eq11} 
x^{k+1}= \frac {x_t+x_{t-1}}{2} -  \text{sign} {(z_t-z_{t-1})} \frac {{|z_t-z_{t-1}|}^N}{2r \mu ^N}, v(x_{t-1})=v(x_t).
\end{equation}

\textit{Stop criterion} of the algorithm, according to which the execution of the algorithm is terminated for current global optimization problem (\ref{eq2}) is the achievement of a predefined accuracy of the problem solution i. e. if the length of the interval $(x_{t-1},x_t)$ containing current point $x^{k+1}$ is less or equal to the predefined accuracy $\Delta _t \leq \varepsilon$ where $t$ is from Rule 5 and $\varepsilon >0$ is the predefined accuracy.

@@

\subsubsection{The fifth approach.} To utilize the search information and the suggestions on the function form most completely, one can restore the objective function values in the points taking onto account the H{\"o}lder condition.
\paragraph{Step 1.} Find the end points of the crossing of majorant and minorant
\begin{equation}\label{eq18} 
\hat{x}',\hat{x}''=\frac {x'+x''}{2}\pm \text{sign}(z''-z')\cdot \frac {1}{2r} \cdot {\left(\frac {z''-z'}{\mu}\right)}^N,
\end{equation}
\paragraph{Step 2.} Restore the objective function values according to the rule
\begin{equation}\label{eq19} 
\tilde{z}(x)=
  \begin{cases}
    z', & {\quad x \leq \hat{x}',}\\
    \frac {z'+z''}{2}- \frac {r \mu}{2} (\| x - x' \| + \| x - x'' \|), & {\quad \hat{x}' < x < \hat{x}'',}\\
    z'',  & {\quad x \geq \hat{x}''.}
  \end{cases}
\end{equation}
This rule can be interpreted as an average value of the minorant and the majorant for a function satisfying the H{\"o}lder condition. Let us give the description of the modification of the global search algorithm in the case of not everywhere computable function based on this approach.

\subsubsection{Description of the modification.} The algorithm remains the same with an introduction of the following minor additions in the computing formulae in the case of falling into the non-computable search regions:
\begin{enumerate} 
  \item The index $v_i=v(x_i)$ is determined by the formula (\ref{eq12}).
  \item When computing current lower estimate of the H{\"o}lder constant $K$ in formula (\ref{eq7}), use the intervals with two computable points only.
  \item The best current objective function value is determined for different values of the index
\begin{equation}\label{eq20} 
z_v^*=
  \begin{cases}
    -\varepsilon _r, & {\quad v=0 ,}\\
    \min \left\{ \phi (x_i): v(x_i)=1, 1 \leq i \leq k \right\}, & {\quad v=1 ,}
  \end{cases}
\end{equation}
where $\varepsilon _r$ is a parameter of the method.

  \item For each interval $(x_{i-1}, x_i), 1 \leq i \leq k+1$, the computing of the characteristic $R(i)$ is performed according to the following computing formulae.
      \begin{itemize} 
        \item \textit{Case 1.} For the intervals with \textbf{two computable points} or with \textbf{one computable point and one boundary point}, apply the classic rules of global search algorithm
\begin{equation}\label{eq21} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-1}-2z_1^*}{r \mu}, & {\quad  v(x_i)=v(x_{i-1})=1}\\
    2 \Delta _i-4 \frac {(z_i-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=1}\\
    2 \Delta _i-4 \frac {(z_{i-1}-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=-1}
  \end{cases}
\end{equation}
        \item \textit{Case 2.}  For the intervals with \textbf{one boundary point and one non-computable point}, perform the imputation of the values for the non-computable point based on the neighbor computable point and apply the formula
\begin{equation}\label{eq22} 
R(i)=
  \begin{cases}
    2 \Delta _i-4 \frac {(z_{i+1}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=0}\\
    2 \Delta _i-4 \frac {(z_{i-2}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=-1}
  \end{cases}
\end{equation}
        \item \textit{Case 3.} For the intervals with \textbf{one non-computable and one computable point}, perform the imputation of the value for the non-computable point based on the neighbor com-putable point and apply the formula
\begin{equation}\label{eq23} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-2})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-2}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=1}\\
     \Delta _i+\frac {{(z_{i+1}-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_{i+1}+z_{i-1}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=0}
  \end{cases}
\end{equation}
        \item \textit{Case 4.}. For the intervals with \textbf{two non-computable points}, perform the imputation of the values for the non-computable points based on the rule (\ref{eq19}) and apply the formula
\begin{equation}\label{eq24} 
R(i)=\Delta _i+\frac {{(\tilde{z}_i-\tilde{z}_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {\tilde{z}_i+\tilde{z}_{i-1}-2z_1^*}{r \mu}, v(x_{i-1})=v(x_i)=0
\end{equation}
        \item \textit{Case 5.} In the case when any imputation is impossible, use the formula
\begin{equation}\label{eq25} 
R(i)=\Delta _i \cdot {\left( 1-\frac {1}{r} \right)}^N+z_0^*.
\end{equation}
    \end{itemize} 
  \item If several intervals have equal characteristics, prefer the computable interval with the lowest number.
  \item If the best interval appears to be the one with two non-computable points, perform the next trial in the middle point of the interval.
\end{enumerate}

@@

[....rus??? ...] $N = 2,3,4,5$ ...

>>>>

\section{Problem Statement}
In the general form, the global optimization problem can be formulated as follows:
\begin{equation}\label{eq1} 
\phi^*=\phi(y^* )=\min_{y \in D} \phi(y), D=\left\{ y \in R^N: a_i \leq y_i \leq b_i, a,b \in R^N,1 \leq i \leq N \right\},
\end{equation}
where $y=(y_1,y_2,...,y_N)$ is the vector of the varied parameters, $D$ is an $N$-dimensional hypercube, and $N$ is the dimension of the problem being solved.

@@

From our experience in solving the applied optimization problems [refs??], we will assume the total volume of the non-computability regions to be a small fraction of the whole volume of the search domain $D$ (about 10-20\%)\footnote{When testing the algorithm, we have considered also more complex problem statements, which the volume of the non-computability region achieved 50-80\% in. The problems of this kind were solved by the algorithm proposed successfully as well but a large number of trials was required to solve the problem because the method attempted to explore the whole non-computability region. In this connection, in the rules of constructing the problems the experiments for the dimensionality $N=5$ were modified so that the non-computability region were located in the corners of the search domain only.}.

@@

\subsubsection{Global search algorithm (GSA)} At every iteration of the global search a \textit{search trial} is executed. We will call a trial a computing the optimized function value  $\phi (x)$ from (\ref{eq2}). The first trial will be performed in the middle internal point $x^1 \in (0,1)$. The choice of the point $x^{k+1}, k \leq 1$ of the next $(k+1)^\text{th}$ trial is performed on the base of the following rules.

\paragraph{Rule 1} Renumber (by the lower indices) the points $x^i, 1 \leq i \leq k$  of the previous trials in the ascending order by their coordinates i. e.
\begin{equation}\label{eq5} 
0=x_0 < x_1 < ... < x_i < ... < x_k=1
\end{equation}
and juxtapose these ones to the values $z_i=\phi (x_i), 1 \leq i \leq k$ computed in these points and to the indices $v_i=v(x_i)$ defined by the rule
\begin{equation}\label{eq6} 
v_i=v(x_i)=
  \begin{cases}
    -1 & {\quad \text{if } x_i \text{ is a boundary point,}}\\
    1  & {\quad \text{if } x_i \text{ is an internal point,}}
  \end{cases}
\end{equation}
the points $x_0=0$ and $x_{k+1}=1$ are introduced additionally (the values $z_0$ and $z_{k+1}$ are undefined) for convenience of the following notations.

\paragraph{Rule 2} Compute current lower estimate
\begin{equation}\label{eq7} 
\mu = \max\left\{ \frac{|z_i-z_{i-1}|}{\Delta _i}, 2 \leq i \leq k \right\} , \Delta _i= \| x_i -x_{i-1} \| =(x_i-x_{i-1})^{1/N},
\end{equation}
of the unknown H{\"o}lder constant $K$ from (\ref{eq4}) for the reduced function $\phi(x)$ where $r, r>1$ is a parameter of the algorithm; if $\mu=0$ accept $\mu=1$.
\paragraph{Rule 3} Determine the best current value of the objective function
\begin{equation}\label{eq8} 
z^*=\min \left\{ \phi (x_i): 1\leq i \leq k \right\}.
\end{equation}

\paragraph{Rule 4} For each interval $(x_{i-1},x_i),1 \leq i \leq k+1$ compute the value $R(i)$ called a \textit{characteristic} of the interval according to the formula
\begin{equation}\label{eq9} 
R(x_i)=
  \begin{cases}
    \Delta _i+\frac {{(z_i-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-1}-2z^*}{r \mu}, & {\quad  v(x_i)=v(x_{i-1})=1}\\
    2 \Delta _i-4 \frac {(z_i-z^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=1}\\
    2 \Delta _i-4 \frac {(z_{i-1}-z^*)}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=-1}
  \end{cases}
\end{equation}

\paragraph{Rule 5} Select the interval $t$ with the largest value of characteristic $R(i):R(t)= \max_{1 \leq i \leq k+1} R(i)$.

\paragraph{Rule 6} Perform the next trial in the middle point of the interval $(x_{t-1},x_t)$ if the indices of its end points are not the same i. e.
\begin{equation}\label{eq10} 
x^{k+1}=\frac {x_t+x_{t-1}}{2}, v(x_{t-1})\neq v(x_t),
\end{equation}
Otherwise, perform the next trial in the point
\begin{equation}\label{eq11} 
x^{k+1}= \frac {x_t+x_{t-1}}{2} -  \text{sign} {(z_t-z_{t-1})} \frac {{|z_t-z_{t-1}|}^N}{2r \mu ^N}, v(x_{t-1})=v(x_t).
\end{equation}

\paragraph{Stop criterion} of the algorithm, according to which the execution of the algorithm is terminated for current global optimization problem (\ref{eq2}) is the achievement of a predefined accuracy of the problem solution i. e. if the length of the interval $(x_{t-1},x_t)$ containing current point $x^{k+1}$ is less or equal to the predefined accuracy $\Delta _t \leq \varepsilon$ where $t$ is from Rule 5 and $\varepsilon >0$ is the predefined accuracy.

@@

\subsubsection{The fifth approach.} To utilize the search information and the suggestions on the function form most completely, one can restore the objective function values in the points taking onto account the H{\"o}lder condition.
\paragraph{Step 1} Find the end points of the crossing of majorant and minorant
\begin{equation}\label{eq18} 
\hat{x}',\hat{x}''=\frac {x'+x''}{2}\pm \text{sign}(z''-z')\cdot \frac {1}{2r} \cdot {\left(\frac {z''-z'}{\mu}\right)}^N,
\end{equation}
\paragraph{Step 2} Restore the objective function values according to the rule
\begin{equation}\label{eq19} 
\tilde{z}(x)=
  \begin{cases}
    z', & {\quad x \leq \hat{x}',}\\
    \frac {z'+z''}{2}- \frac {r \mu}{2} (\| x - x' \| + \| x - x'' \|), & {\quad \hat{x}' < x < \hat{x}'',}\\
    z'',  & {\quad x \geq \hat{x}''.}
  \end{cases}
\end{equation}
This rule can be interpreted as an average value of the minorant and the majorant for a function satisfying the H{\"o}lder condition. Let us give the description of the modification of the global search algorithm in the case of not everywhere computable function based on this approach.

\subsubsection{Description of the modification} The algorithm remains the same with an introduction of the following minor additions in the computing formulae in the case of falling into the non-computable search regions:
\begin{enumerate} 
  \item The index $v_i=v(x_i)$ is determined by the formula (\ref{eq12}).
  \item When computing current lower estimate of the H{\"o}lder constant $K$ in formula (\ref{eq7}), use the intervals with two computable points only.
  \item The best current objective function value is determined for different values of the index
\begin{equation}\label{eq20} 
z_v^*=
  \begin{cases}
    -\varepsilon _r, & {\quad v=0 ,}\\
    \min \left\{ \phi (x_i): v(x_i)=1, 1 \leq i \leq k \right\}, & {\quad v=1 ,}
  \end{cases}
\end{equation}
where $\varepsilon _r$ is a parameter of the method.

  \item For each interval $(x_{i-1}, x_i), 1 \leq i \leq k+1$, the computing of the characteristic $R(i)$ is performed according to the following computing formulae.

        \paragraph{Case 1} For the intervals with \textbf{two computable points} or with \textbf{one computable point and one boundary point}, apply the classic rules of global search algorithm
\begin{equation}\label{eq21} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-1}-2z_1^*}{r \mu}, & {\quad  v(x_i)=v(x_{i-1})=1}\\
    2 \Delta _i-4 \frac {(z_i-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=1}\\
    2 \Delta _i-4 \frac {(z_{i-1}-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=-1}
  \end{cases}
\end{equation}
        \paragraph{Case 2}   For the intervals with \textbf{one boundary point and one non-computable point}, perform the imputation of the values for the non-computable point based on the neighbor computable point and apply the formula
\begin{equation}\label{eq22} 
R(i)=
  \begin{cases}
    2 \Delta _i-4 \frac {(z_{i+1}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=0}\\
    2 \Delta _i-4 \frac {(z_{i-2}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=-1}
  \end{cases}
\end{equation}
        \paragraph{Case 3} For the intervals with \textbf{one non-computable and one computable point}, perform the imputation of the value for the non-computable point based on the neighbor com-putable point and apply the formula
\begin{equation}\label{eq23} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-2})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-2}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=1}\\
     \Delta _i+\frac {{(z_{i+1}-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_{i+1}+z_{i-1}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=0}
  \end{cases}
\end{equation}
        \paragraph{Case 4} For the intervals with \textbf{two non-computable points}, perform the imputation of the values for the non-computable points based on the rule (\ref{eq19}) and apply the formula
\begin{equation}\label{eq24} 
R(i)=\Delta _i+\frac {{(\tilde{z}_i-\tilde{z}_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {\tilde{z}_i+\tilde{z}_{i-1}-2z_1^*}{r \mu}, v(x_{i-1})=v(x_i)=0
\end{equation}
        \paragraph{Case 5} In the case when any imputation is impossible, use the formula
\begin{equation}\label{eq25} 
R(i)=\Delta _i \cdot {\left( 1-\frac {1}{r} \right)}^N+z_0^*.
\end{equation}
  \item If several intervals have equal characteristics, prefer the computable interval with the lowest number.
  \item If the best interval appears to be the one with two non-computable points, perform the next trial in the middle point of the interval.
\end{enumerate}

@@

[....rus...] $N = 2,3,4,5$ ...
